---
title: "STA3180 Project"
author: "Add Names"
date: "2024-10-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(plotly)
library(car)
```
Inlcuding Continents in Dataset:
```{r}
data <- read_csv("/Users/emilymojtabaee/Downloads/group_3_data.csv")
continents <- read_csv("/Users/emilymojtabaee/Downloads/Countries by continents.csv")
continents <- continents %>% rename(reference_area = Country)
data <- merge(data, continents, by = "reference_area")
```
Statistics by continent (mean, median, sd, max, min):
```{r}
datacont <- data[,-1] %>% group_by(Continent)
variables <- c("education", "employment", "absence_of_legal_discrimination", "access_to_productive_and_financial_assets", "reproductive_discrim", "restricted_civil_liberties", "discrimination_in_the_family")
statstable_bycont <- datacont %>% group_by(Continent) %>% 
  summarise(across(all_of(variables), list(mean = mean,
                                           median = median,
                                           sd = sd,
                                           max = max,
                                           min = min)))
statstable_bycont <- data.frame(statstable_bycont)
statstable_bycont
```
Statistics overall (mean, median, sd, max, min):
```{r}
statstable <- data.frame(Mean = sapply(data[variables], mean),
                         Median = sapply(data[variables], median),
                         `Standard Deviation` = sapply(data[variables], sd),
                         Maximum = sapply(data[variables], max),
                         Minimum = sapply(data[variables], min))
statstable
```
Means of just education and employment by continent (Q1+3):
```{r}
minitable <- datacont %>% group_by(Continent) %>% summarize(`Mean Education` = mean(education),`Mean Employment` = mean(employment))
minitable
```
Means of just education and access to financials by continent (Q2):
```{r}
minitable2 <- datacont %>% group_by(Continent) %>% summarize(`Mean Education` = mean(education),`Mean Access` = mean(access_to_productive_and_financial_assets))
minitable2
```
Correlation Matrix:
```{r}
cormatrix <- as.data.frame(cor(data[,variables]))
cormatrix
````
VIF values:
```{r}
reg_for_vif <- lm(employment ~ education + absence_of_legal_discrimination + access_to_productive_and_financial_assets + reproductive_discrim + restricted_civil_liberties + discrimination_in_the_family, data)
vif(reg_for_vif)
```
# For Questions 1 and 3
Seperate Boxplots:
```{r}
ggplot(data, aes(x = Continent, y = education)) + geom_boxplot() + labs(title = "Women's Education by Continent", y = "Education")
ggplot(data, aes(x = Continent, y = employment)) + geom_boxplot() + labs(title = "Women's Employment by Continent", y = "Employment")
```
Combined Boxplots:
```{r}
ggplot(data) +
  geom_boxplot(aes(x = Continent, y = education, fill = "education")) +
  geom_boxplot(aes(x = Continent, y = employment, fill = "employment")) +
  labs(title = "Boxplots of Education and Employment", y = "Education and Employment", x = "Continent")
```
General Education vs Employment Plot:
```{r}
ggplot(data, aes(x = education, y = employment)) + geom_point() + geom_smooth(method = "lm", se = FALSE) + labs(title = "Education vs Employment")
```
Education vs Employment Plot by Continent:
```{r}
ggplot(data, aes(x = education, y = employment, color = Continent)) + geom_point() + geom_smooth(method = "lm", se = FALSE) + labs(title = "Education vs Employment by Country")
```

# For Question 2
Seperate Boxplots:
```{r}
ggplot(data, aes(x = Continent, y = education)) + geom_boxplot() + labs(title = "Women's Education by Continent", y = "Education")
ggplot(data, aes(x = Continent, y = access_to_productive_and_financial_assets)) + geom_boxplot() + labs(title = "Women's Access to Productive and Financial Assets by Continent", y = "Access to Productive and Financial Assets")
```
Combined Boxplots:
```{r}
boxplot <- ggplot(data) +
  geom_boxplot(aes(x = Continent, y = education, fill = "education")) +
  geom_boxplot(aes(x = Continent, y = access_to_productive_and_financial_assets, fill = "Access to Productive and Financial Assets")) +
  labs(title = "Boxplots of Education and Access to Productive and Financial Assets", y = "Education and Access to Productive and Financial Assets", x = "Continent")
ggplotly(boxplot) %>% layout(boxmode = "group")
```
General Education vs Access Plot:
```{r}
ggplot(data, aes(x = education, y = access_to_productive_and_financial_assets)) + geom_point() + geom_smooth(method = "lm", se = FALSE) + labs(title = "Education vs Access to Productive and Financial Assets")
```
Education vs Access by Continent:
```{r}
ggplot(data, aes(x = education, y = access_to_productive_and_financial_assets, color = Continent)) + geom_point() + geom_smooth(method = "lm", se = FALSE) + labs(title = "Education vs Access to Productive and Financial Assets")
```
